There once was a grid at ol' Carkeek

### 2016-08-13 Manuscript/v0.0/carkeek_grid.tex
push misc changes since last visited (2016-05-13)
formatting
removed irrelevant text

Biodiversity surveillance is being revolutionized by DNA-based detection of organisms from environmental samples. ?(specifically speed and scope of ecological studies). % "The surveillance of biodiversity using DNA-based detection from environmental samples"
While this approach has been used for decades to survey micro-organisms (CITE VENTER), it has more recently become used as a tool for surveying macro-organisms (CITE EARLY EDNA).
The technique is founded on the premises that:
(1) organisms shed DNA-containing material into the environment, and 
(2) the DNA contained in this material breaks down over time, 
which leads to the conclusion that the concentration of DNA should decrease with distance from and time since the presence of its source.
%there should be higher concentrations of DNA close to the organism that shed it

%If DNA never broke down, there would be equal concentrations of DNA from two individuals that passed through the same environment many years apart. Similarly, there would be no relationship between DNA concentration and distance from a source.

For an immobile organism (e.g. fungal hyphae) % effectively 
in a static environmental medium (e.g. soil), %n effectively
the relationship between DNA concentration and distance appears straightforward (CITE ???).
But the relationship may be complicated by a dynamic environment: %One might expect the relationship to break down in 
An immobile tree may shed cells into the air, which may be carried by the wind for some distance.
This has been a cause for concern for ecologists interested in surveying aquatic organisms, 
because the robustness of eDNA surveys for detecting and quantifying organisms depends on the distance that DNA travels from a source. %in aquatic environments 
This distance will be affected by the nature of the environment: 
Material is less likely to be transported (quickly over) great distances in an ice-covered lake 
than along a turbulent shoreline or a fast-moving stream.
One expects more DNA downstream of a fish in a river than upstream, and

% Over two marine habitats, we may KNOW their benthic communities are different, their pelagic/planktonic communities MAY be different, and the DNA community MAY be different.
In marine environments, this relationship is further complicated by the fact 
that many organisms have tiny, highly dispersive early life history stages (gametes and larvae).
%that can travel great distances but are difficult to track.
These individuals will obscure the expected relationship between DNA concentration and adult abundance because they 
shed genetic material into the environment and 
are small enough to be captured directly by environmental sampling protocols.
For example, while a traditional survey of a sessile, intertidal bivalve would 
determine that the most biomass lies along the shore, there may be a greater concentration 
of their genetic material at some distance away during a spawning event, or in 
an ocean current carrying many larvae.

If DNA communities are composed mostly of species known

Despite the potential for discrepancy between communities of benthic organisms and their DNA in the water column above, several studies have reported consistent patterns between the two.

Many researchers are justifiably cautious about the ?(adoption) of this new form of data.
Their apprehension is rooted in the premise that traditional survey approaches are more accurate because the chain of inference between observation and ecological data is usually short: A researcher sees two swans in Lake Hopatcong and infers the lake is occupied by at least 2 swans.
DNA based surveys, on the other hand, consist of a longer chain of inference:
DNA sequences are reported by a sequencing machine, the machine identifies the sequence of products of a polymerase chain reaction (PCR), PCR amplifies pieces of DNA from a purified genomic DNA sample, DNA is purified (extracted) from an environmental sample, environmental samples contain DNA from organisms present, the organisms present are representative of the biological community about which we wish to make inference. ?( reverse order? tie to concrete example (swans of Lake Hopatcong)).
Clearly, this process is more complex than visual surveys, as the relationship between several steps is complex or unknown.
But consider that the processes ?(behind | underlying) other more widely-used ecological survey techniques are similarly complex, such as bird surveys based on song, or visual identification of fungal spores. When alternate survey approaches are impossible or inefficient, we are more willing to accept any available survey data, regardless of the complexity or uncertainty underlying it. (microbiologists have enthusiastically relied on DNA-based surveys for years for this reason, (though yes, they also do not have the problem of disconnect between individual and cell)).

The ability of DNA surveys to make quantitative inference about communities has been touted by some (CITE new fish quantitation paper) and doubted by others (CITE european eelgrass PLOSONE).
For example, a study linking (blah blah blah) concluded that "metabarcoding is powerful, yet blind" (CITE european eelgrass).
Conversely, others have reported strong quantitative and intuitive links between DNA-based and traditional survey methods (CITE Port 2016 MOLECO).
These studies usually rely on simple statistical models to link DNA quantity to some measurable ecosystem property like biomass (but see CITE).
When confronted with data collected in ?(complex ways/studies/whatever), simple models ?(may | often) fail to detect relationships when they exist, or vice versa ?(they are prone to inflated risk of BOTH type I and type II error) (CITE, see Woltman 2012).
For example, (CITE, look for that Gelman paper) have demonstrated that when data are structured in a hierarchical fashion (e.g. test scores of students in schools belonging to districts belonging to states), a low number of replicates at the first level of hierarchy (SEE THE PAPER).
Similarly, (describe hospital/school problems).

Shelton et al. (CITE Shelton 2016) outlined an approach for structuring statistical models of DNA surveys that address these issues.
This framework improved on alternative statistical techniques by explicitly accounting for the ?(hierarchical | nested | multilevel) structure of the study design, which allows error and uncertainty at each level to be ?(explicitly accounted for| modeled | propagated throughout the model).
That study demonstrated an improvement in the estimate of higher-level (e.g. ecological community) quantities when the processes linking them to the data are specified.
As an example, it was shown that incorporation of data about the mismatch between primer and template DNA sequence can improve the estimate of the relative abundance of unique DNA templates input to a PCR.

Here, we apply this framework to a DNA survey of ?(nearshore | coastal) marine habitat.
(TODO add commentary on current dogma surrounding distribution of DNA in well-mixed (marine) habitats).
We document the variability associated with lab based ?(procedures | replication | treatment; i.e. filter+DNA+PCR+seq), and the spatial scale over which DNA communities vary in this habitat.
We ?(show that | tested whether) a taxon's spatial distribution predicts ( the slope of the relationship between distance from shore and DNA abundance ~or~ to what degree DNA abundance is explained by distance from shore for each taxon).
We focus partly on species with known life histories that define their spatial distribution (e.g. shallow water livebearing fishes or sessile intertidal organisms with ?(motile/planktonic/pelagic) larvae or gametes).
For these taxa whose spatial distribution is well-documented and restricted, we calculate the rate of change in space and compare this rate among taxa with similar spatial distributions.
In turn, the distribution of rate of change serves as an estimate of the spatial distribution of DNA in this habitat.

We would love to estimate the minimum distance over which eDNA community differences can be detected.

% We measure 'extra-(corporeal|organismal) DNA
Some authors have cautioned against the use of DNA-based macrobial communities in marine environments because they are subject to dynamic physical forces (CITE).
%If environmental circulation

@@

\subsection*{Organismal Life History Data}
We collated coarse-scale data on life history characteristic for each of the major taxonomic groups recovered, 
including dispersal range of the gametes, larvae, and adults, adult habitat type and selectivity, and adult body size. 
Dispersal range was given as an order-of-magnitude approximation of the scale of dispersal: for example, internally fertilized species were assigned a gamete range of 0 km, while broadcast spawners were assigned a gamete range of 10 km. 
Similarly, adult range size was approximated as 0 km (sessile), 1 km (motile but not pelagic), or 10 km (highly mobile, pelagic). 
Variables were specified as 'multiple' for groups known to span more than 1 magnitude of range size. 
For groups to which sequences were annotated with high confidence, but for which life history strategy is diverse or poorly known (e.g. families in the phylum Nemertea), we used conservative, coarse approximations at a higher taxonomic rank. 
We assessed whether or not marine invertebrate taxa are thought to be present in Puget Sound by checking for their presence in a comprehensive checklist of invertebrates of Puget Sound CITE KOZLOFF. 
These data are available as part of the REFERENCE SUPPLEMENTAL DATA.

>>>>

@@

\subsection*{Organismal Life History Data}
We collated coarse-scale data on life history characteristic for each of the major taxonomic groups recovered,
including dispersal range of the gametes, larvae, and adults, adult habitat type and selectivity, and adult body size.
Dispersal range was given as an order-of-magnitude approximation of the scale of dispersal: for example, internally fertilized species were assigned a gamete range of 0 km, while broadcast spawners were assigned a gamete range of 10 km.
Similarly, adult range size was approximated as 0 km (sessile), 1 km (motile but not pelagic), or 10 km (highly mobile, pelagic).
Variables were specified as 'multiple' for groups known to span more than 1 magnitude of range size.
For groups to which sequences were annotated with high confidence, but for which life history strategy is diverse or poorly known (e.g. families in the phylum Nemertea), we used conservative, coarse approximations at a higher taxonomic rank.
%We assessed whether or not marine invertebrate taxa are thought to be present in Puget Sound by checking for their presence in a comprehensive checklist of invertebrates of Puget Sound CITE KOZLOFF.
These data are available as part of the REFERENCE SUPPLEMENTAL DATA.
